Lossy compression of DNA sequencing quality data

### 2016-04-26 bmc_manuscript/manuscript.tex
Added text describing the economic costs for QVZ and RQS. Other small revisions. Need to confirm if RQS smooths or discards quality values for positions in k-mers outside the chosen Hamming distance of dictionary k-mers.

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\cite{Ochoa:2013rt},
\textsc{rqs}~\cite{DBLP:conf/recomb/YuYB14}, and
\textsc{qvz}~\cite{Malysa01102015}.

@@

The economic costs of file size for compression with
\textsc{QualComp}, in our interpretation, appear to have fixed costs
for storing the cluster representatives, and the marginal costs that
are the product of the number of reads times the cost of representing
the model parameters.

\textcolor{red}{Add brief description of \textsc{rqs} and
  \textsc{qvz}, including economic costs.}

\textsc{qvz} extends the idea of using rate-distortion theory introduced in \textsc{QualComp}, allowing additional distortion metrics besides \textsc{mse} to be minimized with respect to user-specified rate (bits/read).

\textsc{rqs} first constructs a dictionary of commonly occurring $k$-mers from either the sequencing data set they seek to compress or a larger population-sized sequencing data set.
During compression, $k$-mers within reads that have a small Hamming distance from a dictionary $k$-mer are identified.
These positions may correspond to SNPs or sequencing error.
Positions within the $k$-mer that differ from the nearest $k$-mer in the dictionary are marked.
Lastly, the quality values of the remaining positions are smoothed.

@@

We secondarily compress all otherwise compressed data sets using the
Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is the
results of these subsequent compressions that we report. Although
other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78} might be particularly effective for
subsequent compression for some compressed data representations
examined, we not explored such possibilities here, and instead
restrict ourselves to \textsc{bz}ip2 because it is commonly used and
readily available. Similarly, it may very well be possible to obtain
increased secondary compression by reordering (e.g., sorting)
compressed representations (e.g., sets of bin designations, model, or
profiles), but these operations have also not been explored here.

>>>>

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\cite{Ochoa:2013rt}, Read Quality Sparsifier
(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}, and Quality Values Zip
(\textsc{qvz)}~\cite{Malysa01102015}.

@@

\textsc{qvz} extends the idea of using rate-distortion theory
introduced in \textsc{QualComp}, allowing distortion metrics other
than \textsc{mse} to be minimized with respect to user-specified rate
(bits/read).

The economic costs of file size for compression with \textsc{QualComp}
or \textsc{qvz}, in our interpretation, appear to be the fixed costs
for storing the cluster representatives, and the marginal costs that
are the product of the number of reads times the cost of representing
the model parameters.

\textsc{rqs} first constructs a dictionary of commonly occurring
$k$-mers from either the sequencing data set they seek to compress or
a larger population-sized sequencing data set. During compression,
$k$-mers within reads that have a small Hamming distance from a
dictionary $k$-mer are identified. The explicit assumption is that
positions differing from the $k$-mers in the dictionary more likely
correspond to SNPs or sequencing errors. These positions that differ
from the nearest $k$-mer in the dictionary are marked. Lastly, the
quality values of the remaining positions are smoothed (discarded? -
see paper).

The economic costs of file size for compression with \textsc{rqs}, in
our interpretation, appear to be the fixed costs of the dictionary of
commonly occurring $k$-mers, and the marginal costs that are the
product of the number of positions of $k$-mers within the chosen
Hamming distance from dictionary $k$-mers times the cost of
representing these positions. It should be noted that the fixed costs
of the dictionary of commonly occurring $k$-mers can be amortized over
reads for which the dictionary may be appropriately used (e.g., reads
from the same species).

@@

We secondarily compress all otherwise compressed data sets using the
Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is the
results of these subsequent compressions that we report. Although
other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78} might be particularly effective for
subsequent compression for some compressed data representations
examined, we not explored such possibilities here, and instead
restrict ourselves to \textsc{bz}ip2 because it is commonly used and
readily available. Similarly, it may very well be possible to obtain
increased secondary compression by reordering (e.g., sorting)
compressed representations (e.g., sets of bin designations, models, or
profiles), but these operations have also not been explored here.
